---
title: "Data Cleaning Tutorial: Matching Dirty Data"
author: "Mark van der Loo"
output:
  beamer_presentation:
    latex_engine: xelatex
    includes:
      in_header: ../tex/header.tex
classoption: "aspectratio=169"
---

## Try the code together with your neighbour

\centering\huge
`02input/merge_with_backbone.R`


## String distance

### Default (Optimal String Alignment distance)

Count number of character deletions, insersions, substitutions and 
transpositions (of adjacent characters)
```{r}
library(stringdist)
stringdist("Ross Ihaka", "Robert Gentleman")
```


## Exact Matching with `match`

```{r}
lookup <- c("Alice","Bob","Carol","Danny")
raw    <- c("Bob","Carl","Rob","bob","Dan","Alice")
i <- match(raw, lookup)
data.frame(raw=raw, matched=lookup[i])
```

## Approximate Matching with `stringdist::amatch`

```{r}
library(stringdist)
j <- amatch(raw, lookup, maxDist=2)
data.frame(raw=raw, matched=lookup[i], amatched=lookup[j])
```

$\to$ Match with closest match, and distance $\leq 2$.

## Optimal string alignment?

```{r}
stringdist("Robert Gentleman", "Gentleman, Robert")
stringdist("Robert Gentleman", "Ross Ihaka")
```

$\to$ OSA wil give a false match (if we allow `maxDist` of 12)


## Alternative: cosine distance 

```{r}
stringdist("Robert Gentleman","Gentleman, Robert"
            , method="cosine", q=2)
stringdist("Robert Gentleman","Ross Ihaka"
            , method="cosine", q=2)
```

### Notes
- Based on counting co-occurrence of character $q$-grams (here: pairs).
- Always between 0 and 1



## More on `amatch`

```{r, eval=FALSE}
amatch(x, table, method, maxDist,...)
```

\begin{tabular}{ll}
\texttt{x}      & character data to be matched\\
\texttt{table}  & the lookup table with clean values\\
\texttt{method} & string distance type\\
\texttt{maxDist}& Maximum distance allowed (depends on \texttt{"method"}!)\\
\texttt{...}    & Extra options depending on \texttt{"method"}
\end{tabular}

### Example

```{r,eval=FALSE}
amatch(raw, lookup, method="cosine", maxDist=0.5, q=3)
```



## Assignment

Merge data from the `companies` dataset with data from `backbone.csv`.

- Using approximate matching on the `"name"` and `"company"` column.
- Think about and try different distance functions and `maxDist`
- Keep your best solution
- Remove rows that cannot be matched
- Write to `02input/myinput.csv`









