---
title: "Data Management and Data Cleaning for Scientists I"
author: "Mark van der Loo"
output:
  beamer_presentation:
    latex_engine: xelatex
    includes:
      in_header: ../tex/header.tex
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Contents


| Lecture | Topics                                       |
|---------|----------------------------------------------|
| 1       | Structuring data and analyses                |
| 2       | Reproducibility and introduction to R        |
| 3       | Data cleaning 1: raw data, data validation   |
| 4       | Data cleaning 2: fixing errors, missing data |



## Materials for these lectures

\centering\huge `github.com/markvanderloo/UFPEL2019`

\begin{picture}(0,0)
\put(100,-80){\includegraphics[height=3cm]{fig/ufpel2019QR.png}}
\end{picture}


# The faces of data

## What do we mean when we say 'data'?

\centering\huge
Data are a representation of information.


##  What a user of data wants

\begin{center}
  \begin{tabular}{rl}
    \multicolumn{2}{l}{CO$_2$ emission (fictional)}\\
    \hline
    \textbf{fuel}    & \textbf{emission}\\
    \hline
    \multicolumn{1}{l}{Petrol}      & 215 \\
    -- of which bio                  & \phantom{0}75  \\
    \multicolumn{1}{l}{Diesel}     & 456 \\
    -- of which bio                  & \phantom{0}89  \\ 
    \hline
  \end{tabular}
\end{center}


## Example: IBGE

\centering
\includegraphics[height=6cm]{fig/ibge.png}

\tiny Source: https://www.ibge.gov.br/en/statistics/economic/industry-and-construction/22733-demography-of-enterprises-and-statistics-of-entrepreneurship.html?=&t=resultados


## What an analyst wants

\begin{center}
  \begin{tabular}{rl}
    \hline
    \textbf{fuel}    & \textbf{emission}\\
    \hline
    \multicolumn{1}{l}{Petrol}      & 215 \\
    -- of which bio                  & \phantom{0}75  \\
    \multicolumn{1}{l}{Diesel}     & 456 \\
    -- of which bio                  & \phantom{0}89  \\ 
    \hline
  \end{tabular}
$\longrightarrow$
  \begin{tabular}{rrl}
    \hline
    \textbf{fuel}      & \textbf{type}  & \textbf{emission}\\
    \hline
    Petrol & regular  & 140 \\
    Petrol & bio      & \phantom{0}75\\
    Diesel& regular  & 367 \\
    Diesel& bio      & \phantom{0}89\\
    \hline
  \end{tabular}
\end{center}


##  What a web developer wants


```{}
        [{"fuel":"petrol","type":"regular","emission":140}, 
         {"fuel":"petrol","type":"bio","emission":75}, 
         {"fuel":"diesel","type":"regular","emission":367}, 
         {"fuel":"diesel","type":"bio","emission":89}]
```

## Example: IBGE

\small
```{}
[{"id":"1501","nome":"Belém","nivel":{"id":"7","nome":"Região metropolitana"}},
{"id":"2301","nome":"Fortaleza","nivel":{"id":"7","nome":"Região metropolitana"}},
{"id":"2601","nome":"Recife","nivel":{"id":"7","nome":"Região metropolitana"}},
{"id":"2901","nome":"Salvador","nivel":{"id":"7","nome":"Região metropolitana"}},
{"id":"3101","nome":"Belo Horizonte","nivel":{"id":"7","nome":"Região metropolitana"}},
{"id":"3301","nome":"Rio de Janeiro","nivel":{"id":"7","nome":"Região metropolitana"}},
{"id":"3501","nome":"São Paulo","nivel":{"id":"7","nome":"Região metropolitana"}},
{"id":"4101","nome":"Curitiba","nivel":{"id":"7","nome":"Região metropolitana"}},
{"id":"4301","nome":"Porto Alegre","nivel":{"id":"7","nome":"Região metropolitana"}}]
```

\normalfont

\hfill \tiny Source: https://servicodados.ibge.gov.br/api/v3/agregados/1705/localidades/N7



##  

```{}
        [{"fuel":"petrol","type":"regular","emission":140}, 
         {"fuel":"petrol","type":"bio","emission":75}, 
         {"fuel":"diesel","type":"regular","emission":367}, 
         {"fuel":"diesel","type":"bio","emission":89}]
```

\begin{center}
$\downarrow$

  \begin{tabular}{rrl}
    \hline
    \textbf{fuel}      & \textbf{type}  & \textbf{emission}\\
    \hline
    Petrol & regular  & 140 \\
    Petrol & bio      & \phantom{0}75\\
    Diesel& regular  & 367 \\
    Diesel& bio      & \phantom{0}89\\
    \hline
  \end{tabular}
\end{center}

## What a database designer sees

\begin{equation*}
\begin{tabular}{ll}
\multicolumn{2}{l}{Fuel}\\
\hline
\textbf{id} & \textbf{name}\\
\hline
11 & petrol  \\
12 & diesel \\
\hline
\end{tabular}\textrm{, }\:
%
\begin{tabular}{ll}
\multicolumn{2}{l}{Type}\\
\hline
\textbf{id} & \textbf{name}\\
\hline
1 & regular \\
2 & bio \\
\hline
\end{tabular}\textrm{, }\:
%
\begin{tabular}{llll}
\multicolumn{3}{l}{Emission}\\
\hline
\textbf{id} & \textbf{fuel} & \textbf{type} & \textbf{amount} \\
\hline
120 & 11 &1 &140\\
121 & 11 &2 &\phantom{0}75\\
123 & 12 &1 &367\\
124 & 12 &2 &\phantom{0}89\\
\hline
\end{tabular}
\end{equation*}


## 

\begin{equation*}
\begin{tabular}{ll}
\multicolumn{2}{l}{Fuel}\\
\hline
\textbf{id} & \textbf{name}\\
\hline
11 & petrol  \\
12 & diesel \\
\hline
\end{tabular}\textrm{, }\:
%
\begin{tabular}{ll}
\multicolumn{2}{l}{Type}\\
\hline
\textbf{id} & \textbf{name}\\
\hline
1 & regular \\
2 & bio \\
\hline
\end{tabular}\textrm{, }\:
%
\begin{tabular}{llll}
\multicolumn{3}{l}{Emission}\\
\hline
\textbf{id} & \textbf{fuel} & \textbf{type} & \textbf{amount} \\
\hline
120 & 11 &1 &140\\
121 & 11 &2 &\phantom{0}75\\
123 & 12 &1 &367\\
124 & 12 &2 &\phantom{0}89\\
\hline
\end{tabular}
\end{equation*}
\begin{center}
$\downarrow$

  \begin{tabular}{rrl}
    \hline
    \textbf{fuel}      & \textbf{type}  & \textbf{emission}\\
    \hline
    Petrol & regular  & 140 \\
    Petrol & bio      & \phantom{0}75\\
    Diesel& regular  & 367 \\
    Diesel& bio      & \phantom{0}89\\
    \hline
  \end{tabular}
\end{center}



## Summarizing

### Presentation

- Convey a (single) message
- Human-readable

### Analyses

- Reusable for (interactive) analyses
- Machine-readable, easy to manipulate

### Transport

- Machine-readable
- Generic, language-independent format

### Storage

- Create, Read, Update, Delete (CRUD)
- Consistent, without (implicit) duplicate information


\begin{picture}(0,0)
\put(250,100){\includegraphics{fig/dataforms.pdf}}
\end{picture}


## Examples of tools

\begin{picture}(0,0)
\put(130,-70){\includegraphics{fig/dataforms.pdf}}

\put(230, -85){\includegraphics[height=5mm]{fig/json.png}}
\put(245, -85){\includegraphics[height=5mm]{fig/xml.png}}

\put(150,-85){\includegraphics[height=5mm]{fig/sql.png}}
\put(125,-85){\includegraphics[height=5mm]{fig/hadoop.png}}

\put(150,35){\includegraphics[height=5mm]{fig/rlogo.png}}
\put(125,35){\includegraphics[height=2mm]{fig/stata.png}}
\put(125,20){\includegraphics[height=5mm]{fig/spss.png}}
\put(125,45){\includegraphics[height=3mm]{fig/sas.png}}

\put(220, 30){\includegraphics[height=5mm]{fig/excel.png}}
\put(245, 25){\includegraphics[height=7mm]{fig/tableau.png}}

\end{picture}


## Why choosing the right tool is important (NYT, 2013)

\begin{center}
\includegraphics[height=0.7\textheight]{fig/excel2.png}
\end{center}

\note{
Reinhart-Rogoff error
}

## Why choosing the right tool is important (Nature, 2019)

\begin{center}
\includegraphics[height=0.7\textheight]{fig/excel3.png}
\end{center}



## Why choosing the right tool is important (Science, 2016)

\begin{center}
\includegraphics[height=0.7\textheight]{fig/excel1.png}
\end{center}

\hfill\tiny Ziemann, Eren, El-Osta \textit{Genome Biology} \textbf{17} 177 (2016)\phantom{MMMMMMMMMMMM}


## Homework assignment

\centering
{\huge Google `excel disasters'}

and spend 30 minutes reading what you find


## Presentation versus analysis

### Spreadsheet software is unsuited for analysis because

- Autoformatting
- Does not force consistency
- Hard to analyze: code is hidden 
- Hard to test
- You see a 'state' not the process


## A bit of terminology

\begin{center}
\begin{tabular}{r@{$\quad\leftrightarrow\quad$}l}
\textbf{Computer scientists} & \textbf{Statisticians}\\
Entity type & Population\\
Entity      & Population unit\\
Attribute   & (Stochastic) variable \\
Value       & Value\\
\end{tabular}
\end{center}


## How to recognize whether data is suited for analysis\footnote{Only for simple rectangular data sets}

### Boxes to tick

1. Does each row correspond to one entity?
2. Are all entities of the same type?
3. Is every entity represented only once?
4. Does every column correspond to a single property for each entity?
5. Are all elements of each column of the same and the correct type?
6. Is the data valid?

### Rule of thumb

Can you make meaningful summary statistics over each column?


## Quizz (1): Ready for analyes?


|            | Alice  | Bob   |  Carol |
|------------|--------|-------|--------|
| Shoe size  |  38    | 43    | 41     |
| Income     |  3300  | 2800  | 4000   |



## Quizz (2): Ready for analyses?

|      | Shoe size | Income |
|------|-----------|--------|
|Alice | 38        | 3300   |
|Bob   | 43        | 2800   |
|Carol | 41        | 4000   |


## Quizz (3): Ready for analyses?

**Income distribution**

|           |€14k -- €20k | €20k -- €40k| €40k -- €80k |  €80K+ | 
|-----------|-------------|-------------|--------------|--------|
| Amsterdam | 20%         |    40%      |    35%       |  5%    |
| Rotterdam | 30%         |    30%      |    38%       |  2%    |
| Den Haag  | 25%         |    35%      |    30%       | 10%    |


## Quizz (4): Ready for analyses?

|      | Age | has job |
|------|-----|---------|
| Dave | 36  | No      |
| Eve  | 5   | Yes     |


## Quizz (5): Ready for analyses?

|             | costs | profit | 
|-------------|-------|--------|
| Retailers   | 50    | 10     |
| Wholesalers | 20    |  5     |
| Total       | 70    | 15     |




## Not all data is 'simple rectangular'

\begin{picture}(0,0)
\put(0,0){\includegraphics[height=3cm]{fig/nsnet.pdf}}
\put(100,-25){\includegraphics[height=3cm]{fig/timeseries.png}}
\put(200,-50){\includegraphics[height=3cm]{fig/bdichtheid.png}}
\put(300,-75){\includegraphics[height=2cm]{fig/image.jpg}}

\put(0,-95){\begin{minipage}{0.8\textwidth}Each data type consists of particular basic elements and is
manipulated\\ with particular basic operations.\end{minipage}}
\end{picture}


# The Statistical Value Chain


## Value Chains

### Porter's value chain (1985)

_The idea of the value chain is based on the process view of organizations, the idea of seeing a manufacturing (or service) organization as a system, made up of subsystems each with inputs, transformation processes and outputs._


## Statistical Value Chain

\begin{center}
\includegraphics[height=3cm]{fig/valuechain.pdf}
\end{center}


### Notes

- This part only pertains to the data processing stage. Collection, design of
experiments, dissemination of results, and so on are not included.
- The fixed points (half-fabricates) are well-defined statistical products.


## Raw data

\begin{center}
\includegraphics[height=15mm]{fig/vc_raw.pdf}
\end{center}

### Your most valueable resource!

- Hard/expensive to obtain
- Keep unchanged
- Backup according to 3-2-1 principle

\begin{picture}(0,0)
\put(250,-20){\includegraphics[height=3cm]{fig/321.png}}
\end{picture}



## The importance of backups



\begin{columns}
  \begin{column}{0.5\textwidth}
    \includegraphics[height=5cm]{fig/hd_failure.png}
    \tiny backblaze.com/blog/backblaze-hard-drive-stats-q2-2019/
  \end{column}
  \begin{column}{0.5\textwidth}
    \begin{block}{Optimistically:}
      \begin{itemize}
       \item On average $P(failure)\approx 0.001$/year.
       \item UFPEL has $\approx 20\,000$ students.
       \item We expect $\approx 20$ failures/year. 
    \end{itemize}
    \end{block}
  \end{column}
\end{columns}



## Input data

\begin{center}
\includegraphics[height=15mm]{fig/vc_input.pdf}
\end{center}

### Technically 'clean' data

- File type is known and can be read
- Data structured for analyses
- Variables are of correct type (number/date/text/categorical$\ldots$)
- Records identified with statistical objects
- Variables identified with statistical properties

### Rule of thumb

You can read this data into your favorite analyses tool,
without errors, with a single expression.

## From raw to input, an example from the LATTES system

\centering
\includegraphics[height=7cm]{fig/tatiana.png}


## From raw to input, an example from the LATTES system

\centering
\includegraphics[height=4cm]{fig/xmlbutton.png}





## From raw to input, an example from the LATTES system

\tiny{}

```{xml, echo=TRUE}
<?xml version="1.0" encoding="ISO-8859-1" standalone="no"?><CURRICULO-VITAE SISTEMA-ORIGEM-XML="LATTES_OFFLINE" NUMERO-IDENTIF
ICADOR="6217846985830016" DATA-ATUALIZACAO="15102019" HORA-ATUALIZACAO="140645"><DADOS-GERAIS NOME-COMPLETO="Tatiana Pereira C
enci" NOME-EM-CITACOES-BIBLIOGRAFICAS="PEREIRA-CENCI,T.;PEREIRA, T.;Pereira-Cenci, Tatiana;PEREIRA, T;PEREIRACENCI, T;Pereira, 
Tatiana;Pereira-Cenci, T.;MULO PATIAS, MAURO ELIAS MESKO" NACIONALIDADE="B" PAIS-DE-NASCIMENTO="Brasil" UF-NASCIMENTO="SP" CID
ADE-NASCIMENTO="Santos" PERMISSAO-DE-DIVULGACAO="NAO" DATA-FALECIMENTO="" SIGLA-PAIS-NACIONALIDADE="BRA" PAIS-DE-NACIONALIDADE
="Brasil"><RESUMO-CV TEXTO-RESUMO-CV-RH="Professora Associada, Departamento de Odontologia Restauradora da Faculdade de Odonto
logia da Universidade Federal de Pelotas. Graduada em Odontologia pela FOB/USP (2001), Especialista em Prótese Dentária pelo H
RAC/USP (2004), Mestre (2006) e Doutora (2008) em Clínica Odontológica/UNICAMP, com PDEE-CAPES na ACTA/Holanda (Nov/2006 a Out
/2007). Tem experiência nos seguintes temas: biofilme, ensaios clínicos randomizados e revisões sistemáticas. É membro do The 
BRIGHTER (Bias, Reporting, Implementation, Guidance, ETHics, IntEgrity and Reproducibility in Research) Meta-Research Group In
itiative." TEXTO-RESUMO-CV-RH-EN= ....
```

\normalfont{}

## LATTES XML format: nodes and attributes

```{xml, echo=TRUE}
<CURRICULO-VITAE SISTEMA-ORIGEM-XML="LATTES_OFFLINE" ...>
   |                 |                   |
  NODE NAME       ATTRIBUTE NAME       ATTRIBUTE VALUE

<DADOS-GERAIS NOME-COMPLETO="Tatiana Pereira Cenci" ...>
   |                 |                   |
  NODE NAME       ATTRIBUTE NAME       ATTRIBUTE VALUE
```



## Structuring `XML` data using R

```{r, echo=TRUE}
library(xml2)
xml  <- read_xml("curriculo.xml")
node <- xml_find_first(xml,"/CURRICULO-VITAE")

d <- data.frame(
    LattesId  = xml_attr(node, "NUMERO-IDENTIFICADOR")
  , Updated   = xml_attr(node, "DATA-ATUALIZACAO")
)
print(d)
```

## Structuring `XML` data using R

```{r, echo=TRUE, message=FALSE}
library(lubridate)

# convert from text to proper data-time format
d$Updated <- dmy(d$Updated)

print(d)

# export to CSV format
write.csv(d, file="CV.csv")
```

## With a little more work

```{r, echo=TRUE, message=FALSE}
node <- xml_find_first(xml, "//ENDERECO-PROFISSIONAL")
d$Institute <- xml_attr(node, "NOME-INSTITUICAO-EMPRESA")
d$Faculty   <- xml_attr(node, "NOME-ORGAO")

nodes <- xml_find_all(xml, "//ARTIGO-PUBLICADO")
d$Articles <- length(nodes)
nodes <- xml_find_all(xml, "//CAPITULO-DE-LIVRO-PUBLICADO")
d$BookChapters <- length(nodes)

print(d)
```


## Valid data

\begin{center}
\includegraphics[height=15mm]{fig/vc_valid.pdf}
\end{center}


### Satisfy domain knowledge constraints

- The last update can not be in the future
- Full professorship under 24 is highly unlikely
- More than $n$ papers/year is unlikely (depending on field)
- $...$
    
### Justification

Invalid data leads to invalid statistical results. 




## Statistics


\begin{center}
\includegraphics[height=15mm]{fig/vc_statistics.pdf}
\end{center}

###  Target output values (aggregates) 

- The resulting numbers for publication

### Note

- These also need to satisfy domain knowledge constraints.



## Output

\begin{center}
\includegraphics[height=15mm]{fig/vc_format.pdf}
\end{center}

### Your paper!

- Formatted, annotated
- Data possibly treated with anonymization techniques (SDC = statistical
  disclosure control)



## The SVC: Remarks

\begin{center}
\includegraphics[height=30mm]{fig/valuechain.pdf}
\end{center}

- Actual data processing is not linear, you will go round a few times.
    - Build up the SVC as your research project progresses.
- Add or remove stages as needed.
- This general idea scales really well.



## Quizz (1)

\begin{center}
\includegraphics[height=2.5cm]{fig/valuechain.pdf}
\end{center}

### Where does the following activity take place?

Formatting date-time variable to `ISO8106` format.

## Quizz (2)

\begin{center}
\includegraphics[height=2.5cm]{fig/valuechain.pdf}
\end{center}

### Where does the following activity take place?

Estimating effect of internationalization academic output.



## Quizz (3)

\begin{center}
\includegraphics[height=2.5cm]{fig/valuechain.pdf}
\end{center}

### Where does the following activity take place?

Standardizing miss-spelled categories, e.g.

- \texttt{"Sim"},\texttt{"si"} $\to$ \texttt{"sim"}
- \texttt{"NO"}, \texttt{"Nao"} $\to$ \texttt{"n\~ao"}

## Quizz (4)

\begin{center}
\includegraphics[height=2.5cm]{fig/valuechain.pdf}
\end{center}

### Where does the following activity take place?

Removing of fixing records where unemployed persons
have a positive income from employment.



## Quizz (5)

\begin{center}
\includegraphics[height=2.5cm]{fig/valuechain.pdf}
\end{center}

### Where does the following activity take place?

Join data with a backbone using probabilistic 
linkage, based on approximate matches between various
columns of the data and the backbone.



## Implementation

\centering\huge

Demo


## Summary

1. Data represents information
2. It is important to choose a representation that suites analyses.
3. Obtaining, cleaning, analyzing data and reporting on results follow a value
   chain structure. It is useful to separate tasks accordingly.

###

There are free and open source tools supporting all necessary methods and
transformations.
   


## For the next lecture: please install R and RStudio

\centering\huge Instructions: `github.com/markvanderloo/UFPEL2019`

\begin{picture}(0,0)
\put(100,-80){\includegraphics[height=3cm]{fig/ufpel2019QR.png}}
\end{picture}





