---
title: "Reproducible Research"
author: "Mark van der Loo"
output:
  beamer_presentation:
    latex_engine: xelatex
    includes:
      in_header: ../tex/header.tex
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Online materials

\centering\huge `github.com/markvanderloo/UFPEL2019`

\begin{picture}(0,0)
\put(100,-80){\includegraphics[height=3cm]{../lecture_1/fig/ufpel2019QR.png}}
\end{picture}

## The reproducibility crisis

\centering
\includegraphics[height=0.7\textheight]{fig/news.jpg}

## Homework (if you want to know more)

\centering\huge
Read the wikipedia article on 'replication crisis'


## ASTM definitions (in engineering)

### Replicability

A measurement is _replicated_ if it is performed in the same
laboratory, by the same operator, using the same apparatus,
within a short time span.


### Reproducibility

A measurement may be _reproduced_ in a different laboratory,
by a different operator, using a different apparatus after an
undetermined time span.


- These terms define how precision and accuracy generalize when relaxing
  certain conditions.


## In official statistics


### UN principles governing international statistical activities

``[...] methods and procedures employed in the production of international
statistics are chosen to meet professional **scientific standards** and are made
transparent for the users.''

### European Statistics Code of Practice

``Based on **scientific principles and methods**, the European Statistical System will
offer and continuously improve a programme of harmonised European statistics''


- No exlicit reference to reproducibility, nor defined precisely, but implied
reproducibility by reference to the scientific method.



## Reproducibility and the scientific method

### Scientific objectivity

Reproducibility  is an aspect of what is called _process objectivity_ of
science. It adheres to the idea that scientific statements are objective in the
sense that they are produced in a way that ``neither depend on contingent social
and ethical values, nor on the individual bias of a
scientist''\footnote{Stanford encyclopedia of philosophy}\footnote{
Definition of concepts is not part of this process}.

### Notes

- Statistical inference quantifies the measure of reproducibility of statistical (sampling) experiments.
- Reproducible data analyses is a very basic requirement underlying reproducibility which in practice turns out to be non-trivial.



## Practical reasons for reproducibility

### 

- Accountability, transparancy
- Efficiency 
- Knowledge sharing
- Process quality
    - Usability (understandable)
    - Maintainability (testable, analyzable)



# Reproducibility of statistical experiments

## Statistical inference


###

> Given the result of my randomized experiment, how do I expect this result to vary if I repeat the experiment?


### Examples

- $p$-value: estimated probability that I will see a different result when the experiment is repeated.
- $95\%$-confidence interval: estimated interval where I expect to see the result when the experiment is repeated.

### Note

$p$ values, confidence intervals, etc express an estimate of the amount of
evidence gatherd by an experiment for a given result. It does _not_ say whether
the result is important.


## Statistical inference

### 1. Estimate a population parameter, based on a randomized sample


### 2. Analytical route

a. Derive a distribution of the estimate, in the limit of sampling over and over again.
b. Based on a your actual sample, estimate the parameters of the distribution of your estimate over repeated samples.



### 2. Computational (bootstrap) route

a. From your randomized sample, draw a new sample (SRSWR), and estimate.
b. Repeat step 2 many times to estimate the distribution of your estimate over repeated samples.


### 3. Use estimated sampling distribution for $p$ values, CI, $\ldots$


## Example: estimate mean turnover from a sample of retailers

```{r, fig.height=5}
uured <- "#F52A01"
invd <- function(x,mu,sd){
  sqrt(-2*sd*sd*log(sqrt(2*pi)*sd*dnorm(x,mu,sd))) + mu
}
library(validate)
data(retailers)
to <- with(retailers,{
  ifelse(turnover/staff > 10000,turnover/1000,turnover)
})
par(cex=1.3)
to <- to[!is.na(to)]
sdhat <- sd(to,na.rm=TRUE)
muhat <- mean(to,na.rm=TRUE)
n <- length(to)
x <- seq(min(to),max(to),length.out=300)

hist(to,breaks=15,main='Turnover of 51 retailers',xlab="turnover (kEUR)",las=1,freq=TRUE)
d <- dnorm(x,muhat, sdhat/sqrt(length(to)))*10000
lines(x,d,lwd=2,col=uured)

left <- qnorm(0.025,mean=muhat,sd = sdhat/sqrt(n))
right <- qnorm(0.975,mean=muhat,sd=sdhat/sqrt(n))
i <- x<=right & x >= left
xpgn <- c(range(x[i]),rev(x[i]))
ypgn <- c(0,0,rev(d[i]))
polygon(xpgn,ypgn,col=uured,density = 7,lwd=2)
abline(v=muhat,lwd=2)
#text(2700,13.2, bquote(hat(theta) == .(signif(muhat,4))))
text(4500,9,substitute( 
  paste(Theta, " ~ N(", hat(theta), "=", m, ", ", hat(sigma)/sqrt(n), "=", s2, ")"),
    list(m = signif(muhat,4), s2 = signif(sdhat/sqrt(length(to)),4))
  ))
legend("right",legend="95% conf. int.", bty="n",density=14,fill=uured,border=uured)
```


## Example 2: Computational inference of windsorized mean

```{r,eval=TRUE, echo=TRUE}
# Mean, windsorized at median + 3 times median absolute deviation
wins_mean <- function(x){
  limit <- median(x) +  3 * mad(x)
  x[x > limit] <- limit
  mean(x)
}
```

## Estimate windsorized mean from sample of retailers

```{r, fig.height=5}
library(boot)
stat <- function(d,i){
  wins_mean(d[i])
}
set.seed(1)
b <- boot(to, statistic = function(d,i) wins_mean(d[i]), R=500)
par(cex=1.3)
hist(to,breaks=15,main='Turnover of 51 retailers',xlab="turnover (kfl)",las=1,freq=TRUE)
d <- density(b$t)
lines(d$x,d$y*1e4,col=uured,lwd=2)
text(5000,10,substitute(
  paste(E[b],hat(theta)," = ", m
        ,", ", hat(sigma)[b]," = ", ss )
  , list(m=signif(mean(b$t),4),ss=signif(sd(b$t),4))
))
abline(v=mean(b$t),lwd=2)   
q <- quantile(b$t,probs=c(0.025,0.975))
i <- d$x > q[1] & d$x < q[2] 
xx <- c(q, rev(d$x[i]))
yy <- c(c(0,0),rev(d$y[i]))
polygon(xx,yy*1e4,density = 7,col=uured,lwd=2)
     
legend("right",legend="95% conf. int.", bty="n",density=14,fill=uured,border=uured)
```



# Reproducibility of data processing



## Scholarship in computational science

_An article about computational science in a scientific
journal is **not** the scholarship itself, it is merely **advertising**
of the scholarschip. The actual scholarship is the complete
software development environment and the complete set of instructions
which generated the figures.\footnote{Buckheit and Donovo (1995)}_




## Main idea of a reproducible analyses

The product of an analyses is not a written article, but rather
a _compendium\footnote{Gentleman and Temple Lang (2004)}_ containing

- The data being analyzed
- Runnable code used to compute the numbers
- The text explaining the problem, methods and interpretation.

### Note

This requires some skills in tools and programming.


## Dynamic documents: literate programming

\begin{center}
\includegraphics[height=0.8\textheight]{fig/knuth-web.pdf}
\end{center}

\tiny{Knuth (1992)}

## Reproducible research using `rmarkdown`


\huge\centering Demo


# Keeping track of changes: version control

## Why version control?

\begin{columns}
\begin{column}{0.3\textwidth}
\includegraphics[height=0.8\textheight]{fig/folder.png}
\end{column}\begin{column}{0.7\textwidth}
\begin{itemize}
\item Keeping track of updates manually is messy, difficult, and depends on discipline.
\item Fortunately this can be automated.
\end{itemize}
\end{column}
\end{columns}


## 

\centering
\includegraphics[height=0.8\textheight]{fig/ph1/UG01.jpg}


## 

\centering
\includegraphics[height=0.8\textheight]{fig/ph1/UG02.jpg}

```{bash, eval=FALSE, echo=TRUE}
$> vi myfile.R
```


## 

\centering
\includegraphics[height=0.8\textheight]{fig/ph1/UG03.jpg}

```{bash, eval=FALSE, echo=TRUE}
$> git add myfile.R
```


## 

\centering
\includegraphics[height=0.8\textheight]{fig/ph1/UG04.jpg}

```{bash, eval=FALSE, echo=TRUE}
$> vi myfile.R
```


## 

\centering
\includegraphics[height=0.8\textheight]{fig/ph1/UG05.jpg}

```{bash, eval=FALSE, echo=TRUE}
$> git add myfile.R
$> git commit -m "fixed a bug"
```


## 

\centering
\includegraphics[height=0.8\textheight]{fig/ph1/UG06.jpg}

```{bash, eval=FALSE, echo=TRUE}
$> vi myfile.R
```


## 

\centering
\includegraphics[height=0.8\textheight]{fig/ph1/UG07.jpg}

```{bash, eval=FALSE, echo=TRUE}
$> vi myfile.R
```


## 

\centering
\includegraphics[height=0.8\textheight]{fig/ph1/UG08.jpg}

```{bash, eval=FALSE, echo=TRUE}
$> vi myfile.R
```


## 

\centering
\includegraphics[height=0.8\textheight]{fig/ph1/UG09.jpg}

```{bash, eval=FALSE, echo=TRUE}
$> git log --oneline
$> git checkout be9055b .
```

## 

\centering
\includegraphics[height=0.8\textheight]{fig/ph1/UG10.jpg}



## 

\centering
\includegraphics[height=0.8\textheight]{fig/ph1/UG11.jpg}


##

\centering\huge Demo

